---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r, load necessary packages, echo=TRUE}

library(dplyr)
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data

```{r,include=TRUE,echo=TRUE}
activity= read.csv("activity.csv")


activity$date= as.Date(activity$date)

#extract day from date then collapse steps sum function grouping by day
activity1= dplyr::mutate(activity,day=day(activity$date))

activity1=group_by(activity1,day)%>% summarise(steps=sum(steps,na.rm = T))

```


## What is mean total number of steps taken per day?

```{r,include=T,echo=TRUE}
ggplot(data = activity1,aes(day,steps)) + geom_histogram(stat = "identity") + labs(title= " total number of steps taken each day",ylab="Steps",xlab="Day") + theme_bw()


```

## What is the average daily activity pattern?
```{r,include=TRUE,echo=TRUE}
#we have already grouped our data by day so we proceed to apply mean and median fun on summarise

group_by(activity1,day)%>%summarise(mean_steps=mean(steps,na.rm = T),median_steps = median(steps,na.rm = T))

```
##Time series plot of the average number of steps taken
```{r,include=T,echo=T}
#group by interval then collapse 

activity2=group_by(activity,interval)%>% summarise(steps=sum(steps,na.rm = T))

ggplot(activity2,aes(x=interval,y=steps))+ geom_line() + labs(title="Time series plot of the average number of steps taken",xlab="5 - minutes interval per day",ylab= "Steps") + theme_bw()


```

##The 5-minute interval that, on average, contains the maximum number of steps
```{r,include=T,echo=T}

summary(activity$steps,na.rm=T)

filter(activity,steps==806)

#the interval with with maximum steps is 615
```


## Imputing missing values

```{r,include=T,echo=T}
#we replace all missing values with mean

activity4= activity

activity4$steps[is.na(activity4$steps)]= as.integer(mean(activity4$steps,na.rm=T))

```

##Histogram of the total number of steps taken each day after missing values are imputed
```{r,include=T,echo=T}
activity4$date= as.Date(activity4$date)

#extract day from date then collapse steps sum function grouping by day
activity4= dplyr::mutate(activity4,day=day(activity4$date))

activity4=group_by(activity4,day)%>% summarise(steps=sum(steps,na.rm = T))

ggplot(data = activity4,aes(day,steps)) + geom_histogram(stat = "identity") + labs(title= " total number of steps taken each day",ylab="Steps",xlab="Day") + theme_bw()

```

## Are there differences in activity patterns between weekdays and weekends?
```{r,include=T,echo=T}

activity5=mutate(activity,weekdays=weekdays(activity$date))

##create two rows and one column panel in plot
par(mfrow = c(2,1))
##weekday dataset
activity5wd= filter(activity5,weekdays %in% c("Friday","Monday","Thursday","Tuesday","Wednesday"))

activity5wdend= filter(activity5,weekdays %in% c("Saturday","Sunday"))

plot(activity5wd$steps, type = "l", main = "weekdays"
     ,xlab = "the 5-minute interval"
     ,ylab = "the average number of steps")

plot(activity5wdend$steps, type = "l", main = "weekdends"
     ,xlab = "the 5-minute interval"
     ,ylab = "the average number of steps")
```

